Notes on robust solvers for systems of coupled multiscale PDEs

### 2019-08-03 solver/problem_gamma.tex
typos

While it can be easily seen that P1-P1 discretization is inf-sup stable for
the problem the pair which we are interested in is P1-P0 as piece-wise constant
elements were used in the previous section. We recall that stable P1-P1 discretization
for $\mathcal{A}_2$ (using different meshes for $\Omega$, $\Gamma$ is also possible,
cf. \cite{burman2009interior}, however, it is not clear how to extend this result
to 3$d$-1$d$ problem \eqref{eq:problem2}.

@@

  Table \ref{tab:stokes} shows that
  iterations are stable when $\epsilon\gg 1$ while they are unbounded for
  $\epsilon<1$. A heuristic explanation for
  the observation can based on Schur complement reasoning. Indeed the
  Schur complement of $\mathcal{A}_h$ consists of two parts corresponding
  to the two Stokes problems
  \[
  S = \epsilon^2\nabla_h\cdot(-\Delta_h)^{-1}\nabla_h + \nabla_h\cdot(-\Delta_h)^{-1}\nabla_h.
  \]
  Owing the discrete inf-sup condition the spectrum of the first part is
  bounded by $\epsilon^2 I_h$ for all $h$. However, for the second component
  a similar bound does not hold (as the corresponding discretization is unstable)
  and this part can become dominant for small $\epsilon$.

@@

\subsection{Stabilized $\mathcal{A}_1$} We construct the stabilizing term
based on two observations. (i) In \S\ref{sec:problem_omega} the stabilization
for $p\in Q_h\subset H^{-1/2}$ in takes the form the facet integral with integrand $h^{2}\jump{p}\jump{q}$.
(ii) For P0 element $-\Delta_h$ is spectrally equivalent to operator involving
facet integrals with integrand $h^{-1}\jump{p}\jump{q}$. In turn we assume
that the stabilization for $H^s$ norms takes the form $h^{p}\jump{p}\jump{q}$
and by interpolation from (i) and (ii) $p=-2s+1$. Thus the stabilizing term for $H^{-1}$ shall
involve the cube of the mesh size and we consider a discrete formulation of \eqref{eq:A1}
in terms of the operator
%
\[
\langle\mathcal{A}_{1, h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
    \rangle
    =
    \int_{\Gamma}\nabla u\cdot\nabla v+uv\,\mathrm{d}x+\int_{\Gamma}uq + vp\,\mathrm{d}x-
    \sum_{F\in\mathcal{F}}h^3\jump{p}\jump{q}\,\mathrm{d}s.
\]
The preconditioner for $\mathcal{A}_{1, h}$ is then the Riesz mapping which
includes the negative stabilized term
\[
\langle\mathcal{B}^{-1}_{1, h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
    \rangle
    =
    \int_{\Gamma}\nabla u\cdot\nabla v+uv\,\mathrm{d}x+
    \langle(-\Delta+I)^{-1}p, q\rangle+
    \sum_{F\in\mathcal{F}}h^3\jump{p}\jump{q}\,\mathrm{d}s.
\]

Table \ref{tab:A1} shows error convergence of the proposed formulation
together with the condition number of the preconditioner problem. We remark
that the manufactured solution on the unit interval has $g=0$ so that $p=f=\cos(\pi x)$
and $u=0$ is the solution. It can be seen that the formulation is stable and
yields converging solutions.

>>>>

While it can be easily seen that P1-P1 discretization is inf-sup stable for
the problem the pair which we are interested in is P1-P0 as piece-wise constant
elements were used in the previous section. We recall that stable P1-P1 discretization
for $\mathcal{A}_2$ (using different meshes for $\Omega$, $\Gamma$ is also possible,
cf. \cite{burman2009interior}), however, it is not clear how to extend this result
to 3$d$-1$d$ problem \eqref{eq:problem2}.

@@

  Table \ref{tab:stokes} shows that
  iterations are stable when $\epsilon\gg 1$ while they are unbounded for
  $\epsilon<1$. A heuristic explanation for
  the observation can be based on Schur complement reasoning. Indeed the
  Schur complement of $\mathcal{A}_h$ consists of two parts corresponding
  to the two Stokes problems
  \[
  S = \epsilon^2\nabla_h\cdot(-\Delta_h)^{-1}\nabla_h + \nabla_h\cdot(-\Delta_h)^{-1}\nabla_h.
  \]
  Owing to the discrete inf-sup condition the spectrum of the first part is
  bounded by $\epsilon^2 I_h$ for all $h$. However, for the second component
  a similar bound does not hold (as the corresponding discretization is unstable)
  and this part can become dominant for small $\epsilon$.

@@

\subsection{Stabilized $\mathcal{A}_1$} We construct the stabilizing term
based on two observations. (i) In \S\ref{sec:problem_omega} the stabilization
for $p\in Q_h\subset H^{-1/2}$ takes the form of the facet integral with integrand $h^{2}\jump{p}\jump{q}$.
(ii) For P0 elements $-\Delta_h$ is spectrally equivalent to operator involving
facet integrals with integrand $h^{-1}\jump{p}\jump{q}$. In turn we assume
that the stabilization for $H^s$ norms takes the form $h^{p}\jump{p}\jump{q}$
and by interpolation from (i) and (ii) $p=-2s+1$. Thus the stabilizing term for $H^{-1}$ shall
involve the cube of the mesh size and we consider a discrete formulation of \eqref{eq:A1}
in terms of the operator
%
\[
\langle\mathcal{A}_{1, h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
    \rangle
    =
    \int_{\Gamma}\nabla u\cdot\nabla v+uv\,\mathrm{d}x+\int_{\Gamma}uq + vp\,\mathrm{d}x-
    \sum_{F\in\mathcal{F}}h^3\jump{p}\jump{q}\,\mathrm{d}s.
\]
The preconditioner for $\mathcal{A}_{1, h}$ is then the Riesz mapping which
includes the negative stabilized term
\[
\langle\mathcal{B}^{-1}_{1, h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
    \rangle
    =
    \int_{\Gamma}\nabla u\cdot\nabla v+uv\,\mathrm{d}x+
    \langle(-\Delta+I)^{-1}p, q\rangle+
    \sum_{F\in\mathcal{F}}h^3\jump{p}\jump{q}\,\mathrm{d}s.
\]

Table \ref{tab:A1} shows error convergence of the proposed formulation
together with the condition number of the preconditioned problem. We remark
that the manufactured solution on the unit interval has $g=0$ so that $p=f=\cos(\pi x)$
and $u=0$ is the solution. It can be seen that the formulation is stable and
yields converging solutions.

### 2019-08-03 solver/problem_omega.tex
typos

With Figure \ref{fig:coupled_domain} in mind we wish to solve the 2$d$-1$d$
coupled problem: Find $u\in V=H^1_{0, \Gamma_{D}}(\Omega)$,
$p\in Q=H^{-1/2}(\Gamma)$ such that
 %
\begin{equation}\label{eq:A2}
\begin{aligned}
  &\int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + &{\int_{\Gamma}p v} &= \int_{\Omega}f v\,\mathrm{d}x\quad\forall v\in V,\\
  &{\int_{\Gamma}q u\,\mathrm{d}s} + &\phantom{\int{\Gamma}p v} &= \int_{\Gamma}g q\,\mathrm{d}s\quad\forall q\in Q.\\
\end{aligned},
\quad\mbox{or equivalently}\quad
\mathcal{A}_2\begin{pmatrix}
u\\
p\\
\end{pmatrix} = L.
\end{equation}
%
It can be seen that $\mathcal{A}_2$ is an isomorphism from $W=V\times Q$
to $W^{'}$ and in turn by \cite{mardal2011preconditioning} operator
\[
\mathcal{B}_2=\text{diag}(-\Delta_{\Omega}, (-\Delta_{\Gamma}+I)^{-1/2})^{-1}
\]
is a cannonical preconditioner for $\mathcal{A}_2$. Considering first the
case (i) from Figure \ref{fig:domains}, i.e. trace mesh of $\Omega_h$ is
$\Gamma_h$ and using (stable) P1-P1 discretization Table \ref{tab:A2} shows
the both the condition number and the number of MinRes iterations of $\mathcal{B}_2\mathcal{A}_2$
are bounded.

In order to arrive at the formulation where the discrete multiplier space
$Q_h$ is setup on the $\Gamma$-intersected cells of $\Omega_h$ let us first
consider an intermediate formulation with a piecewise constant (P0) multiplier defined
on $\Gamma_h$, cf. Figure \ref{fig:domains} (ii) and (iii). It is well known, e.g.
\cite[Ch 11.3]{steinbach2007numerical}, that if trace mesh of $\Omega_h$ is $\Gamma_h$ then P1-P0 discretization is
unstable. However the pair is stable if $\Gamma_h$ is coarser than the trace
mesh or a stabilization is employed. In the application we have in mind
the mesh of $\Omega$ is in general coarser than that of $\Gamma$ and thus
stabilized formulation are of interest. To highlight the difference in resolution
between the meshes we shall use different subscripts, i.e. $\Omega_H$ and
$\Gamma_h$. We remark that stabilization for piecewise linear Lagrange multipliers
(on $\Gamma$) is discussed in \cite{burman2009interior}.

Let $V_H=V_H(\Omega_H)$, $Q_h=Q_h(\Gamma_h)$ be the finite element approximations
of $V$ and $Q$ in terms of P1 and P0 elements respectively. Following \cite{burman2014projection}
the stabilized formulation of \eqref{eq:A2} reads: Find $u \in V_H$ and
$p \in Q_h$ such that
%
\[
\begin{aligned}
  &\int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + &{\int_{\Gamma}p v\,\mathrm{d}s} &= \int_{\Omega}f v\,\mathrm{d}x\quad\forall v\in V_H,\\
  &{\int_{\Gamma}q u\,\mathrm{d}s} + &-\sum_{F\in\mathcal{F}}\int_{F}h^2\jump{p}\jump{q} \mathrm{d}s &= \int_{\Gamma}g q\,\mathrm{d}s\quad\forall q\in Q_h,\\
\end{aligned},
\quad\mbox{or equivalently}\quad
\mathcal{A}_{2, \Gamma_h}\begin{pmatrix}u\\p\\\end{pmatrix} = L.
\]
%
Here $\mathcal{F}$ is the union of internal and $\Gamma_N$-intersecting
facets of $\Gamma_h$. A possible preconditioner for $\mathcal{A}_{2,\Gamma_h}$,
which is based on the mapping properties of the continuous problem is a Riesz
map with respect to the inner product in induced by $\mathcal{B}^{-1}_{2, \Gamma_h}$
\[
\langle
\mathcal{B}^{-1}_{2, \Gamma_h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
\rangle
    =
    \int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + \langle(-\Delta_{\Gamma}+I)^{-1/2}p, q\rangle
    + \sum_{F\in\mathcal{F}}\int_{F}h^2\jump{p}\jump{q} \mathrm{d}s.
\]
Robustness of the preconditioner can be seen in Table \ref{tab:A2} in both
the case where $\Gamma_h$ does not and does intersect the cell interior of $\Omega_h$ (cases
(iii) and (iv) in Figure \ref{fig:domains}). A potential difficulty with
$\mathcal{B}_{2, \Gamma_h}$ is, however, its generalization of the call of $Q_h$ defined
on intersected cells (interiors). In particular, the fractional norm can be
an problematic. To avoid the issue, we therefore consider an alternative
preconditioner based on \cite[\S 4.A]{burman2014projection}
\[
\langle
\tilde{\mathcal{B}}^{-1}_{2, \Gamma_h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
\rangle
    =
    \int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + \int_{\Gamma}h^{-1}u v\,\mathrm{d}s + \int_{\Gamma} h p q\,\mathrm{d}s
    + \sum_{F\in\mathcal{F}}\int_{F}h^2\jump{p}\jump{q} \mathrm{d}s.
\]
Note that the fractional norm on the multiplier has been replaced by the $h$-weighted
$L^2$ norm. Furthermore, there is an additional control of the trace of $u$ on
the curve by $h^{-1}$-weighted $L^2$ norm. We remark that \cite{burman2014projection} proves
the inf-sup condition for $\mathcal{A}_{2, \Gamma_h}$ using the norms on $V_H$ and
$Q_h$ as follows
\[
\norm{u}^2_{V_H} = \norm{\semi{\nabla u}}^2_0 + \int_{\Gamma}h^{-1} u v\,\mathrm{d}s
\quad\mbox{and}\quad
\norm{p}_{Q_h} = \norm{h^{1/2} p}_0.
\]

Robustness of the preconditioner is
shown in Table \ref{tab:A2}; in the cut case (iii) the preconditioner is more
efficinet than $\mathcal{B}_{2, \Gamma_h}$. We remark that the error convergence
of $\norm{u-u_H}_1$ is reduced in this case due to the fact that the kink
of the solution cannot be captured within the element.

Finally, we extend $\mathcal{A}_{2, \Gamma_h}$ to the formulation with Lagrange
mutliplier on the intersected elements. To this end, let $\mathcal{S}_H$ denote
elements of $\Omega_H$ intersected by $\Gamma_h$. The space $Q_H$ now consists
of piecewise constants on $\mathcal{S}_H$. Following \cite[\S 4.B]{burman2014projection}
we consider problem: Find $u\in V_H$, $p\in  Q_H$ such that
%
\[
\begin{aligned}
  &\int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + &{\int_{\Gamma}p v\,\mathrm{d}s} &= \int_{\Omega}f v\,\mathrm{d}x\quad\forall v\in V_H,\\
  &{\int_{\Gamma}q u\,\mathrm{d}s} + &-\sum_{F\in\partial\mathcal{S}}\int_{F}H\jump{p}\jump{q} \mathrm{d}s &= \int_{\Gamma}g q\,\mathrm{d}s\quad\forall q\in Q_H,\\
\end{aligned},
\quad\mbox{or equivalently}\quad
\mathcal{A}_{2, \Omega_H}\begin{pmatrix}u\\p\\\end{pmatrix} = L.
\]
%
Here $\partial\mathcal{S}$ denotes interior and Neumann boundary intersecting
facets of $\mathcal{S}$. We remark that the norms for the inf-sup condition (shown
in \cite{burman2014projection} of $\mathcal{A}_{2, \Omega_H}$ are identical to $\mathcal{A}_{2, \Gamma_g}$.

@@

    formulations of \eqref{eq:A2}, $H_0=2^{-4}$ and $h=H/3$.

>>>>

With Figure \ref{fig:coupled_domain} in mind we wish to solve the 2$d$-1$d$
coupled problem: Find $u\in V=H^1_{0, \Gamma_{D}}(\Omega)$,
$p\in Q=H^{-1/2}(\Gamma)$ such that
 %
\begin{equation}\label{eq:A2}
\begin{aligned}
  &\int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + &{\int_{\Gamma}p v} &= \int_{\Omega}f v\,\mathrm{d}x\quad\forall v\in V,\\
  &{\int_{\Gamma}q u\,\mathrm{d}s} + &\phantom{\int{\Gamma}p v} &= \int_{\Gamma}g q\,\mathrm{d}s\quad\forall q\in Q.\\
\end{aligned},
\quad\mbox{or equivalently}\quad
\mathcal{A}_2\begin{pmatrix}
u\\
p\\
\end{pmatrix} = L.
\end{equation}
%
It can be seen that $\mathcal{A}_2$ is an isomorphism from $W=V\times Q$
to $W^{'}$ and in turn by \cite{mardal2011preconditioning} operator
\[
\mathcal{B}_2=\text{diag}(-\Delta_{\Omega}, (-\Delta_{\Gamma}+I)^{-1/2})^{-1}
\]
is a cannonical preconditioner for $\mathcal{A}_2$. Considering first the
case (i) from Figure \ref{fig:domains}, i.e. trace mesh of $\Omega_h$ is
$\Gamma_h$ and using (stable) P1-P1 discretization Table \ref{tab:A2} shows
that both the condition number and the number of MinRes iterations of $\mathcal{B}_2\mathcal{A}_2$
are bounded.

In order to arrive at the formulation where the discrete multiplier space
$Q_h$ is setup on the $\Gamma$-intersected cells of $\Omega_h$ let us first
consider an intermediate formulation with a piecewise constant (P0) multiplier defined
on $\Gamma_h$, cf. Figure \ref{fig:domains} (ii) and (iii). It is well known, e.g.
\cite[Ch 11.3]{steinbach2007numerical}, that if trace mesh of $\Omega_h$ is $\Gamma_h$ then P1-P0 discretization is
unstable. However the pair is stable if $\Gamma_h$ is coarser than the trace
mesh or a stabilization is employed. In the application we have in mind
the mesh of $\Omega$ is in general coarser than that of $\Gamma$ and thus
stabilized formulation is of interest. To highlight the difference in resolution
between the meshes we shall use different subscripts, i.e. $\Omega_H$ and
$\Gamma_h$. We remark that stabilization for piecewise linear Lagrange multipliers
(on $\Gamma$) is discussed in \cite{burman2009interior}.

Let $V_H=V_H(\Omega_H)$, $Q_h=Q_h(\Gamma_h)$ be the finite element approximations
of $V$ and $Q$ in terms of P1 and P0 elements respectively. Following \cite{burman2014projection}
the stabilized formulation of \eqref{eq:A2} reads: Find $u \in V_H$ and
$p \in Q_h$ such that
%
\[
\begin{aligned}
  &\int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + &{\int_{\Gamma}p v\,\mathrm{d}s} &= \int_{\Omega}f v\,\mathrm{d}x\quad\forall v\in V_H,\\
  &{\int_{\Gamma}q u\,\mathrm{d}s} + &-\sum_{F\in\mathcal{F}}\int_{F}h^2\jump{p}\jump{q} \mathrm{d}s &= \int_{\Gamma}g q\,\mathrm{d}s\quad\forall q\in Q_h,\\
\end{aligned},
\quad\mbox{or equivalently}\quad
\mathcal{A}_{2, \Gamma_h}\begin{pmatrix}u\\p\\\end{pmatrix} = L.
\]
%
Here $\mathcal{F}$ is the union of internal and $\Gamma_N$-intersecting
facets of $\Gamma_h$. A possible preconditioner for $\mathcal{A}_{2,\Gamma_h}$,
which is based on the mapping properties of the continuous problem is a Riesz
map with respect to the inner product in induced by $\mathcal{B}^{-1}_{2, \Gamma_h}$
\[
\langle
\mathcal{B}^{-1}_{2, \Gamma_h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
\rangle
    =
    \int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + \langle(-\Delta_{\Gamma}+I)^{-1/2}p, q\rangle
    + \sum_{F\in\mathcal{F}}\int_{F}h^2\jump{p}\jump{q} \mathrm{d}s.
\]
Robustness of the preconditioner can be seen in Table \ref{tab:A2} in both
the case where $\Gamma_h$ does not and does intersect the cell interior of $\Omega_h$ (cases
(iii) and (iv) in Figure \ref{fig:domains}). A potential difficulty with
$\mathcal{B}_{2, \Gamma_h}$ is, however, its generalization to the case of $Q_h$ defined
on intersected cells (interiors). In particular, the fractional norm can be
problematic. To avoid the issue, we therefore consider an alternative
preconditioner based on \cite[\S 4.A]{burman2014projection}
\[
\langle
\tilde{\mathcal{B}}^{-1}_{2, \Gamma_h}\begin{pmatrix}u\\p\end{pmatrix},
  \begin{pmatrix}v\\q\end{pmatrix}
\rangle
    =
    \int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + \int_{\Gamma}h^{-1}u v\,\mathrm{d}s + \int_{\Gamma} h p q\,\mathrm{d}s
    + \sum_{F\in\mathcal{F}}\int_{F}h^2\jump{p}\jump{q} \mathrm{d}s.
\]
Note that the fractional norm on the multiplier has been replaced by the $h$-weighted
$L^2$ norm. Furthermore, there is an additional control of the trace of $u$ on
the curve by $h^{-1}$-weighted $L^2$ norm. We remark that \cite{burman2014projection} proves
the inf-sup condition for $\mathcal{A}_{2, \Gamma_h}$ using the norms on $V_H$ and
$Q_h$ as follows
\[
\norm{u}^2_{V_H} = \norm{\nabla u}^2_0 + \int_{\Gamma}h^{-1} u v\,\mathrm{d}s
\quad\mbox{and}\quad
\norm{p}_{Q_h} = \norm{h^{1/2} p}_0.
\]

Robustness of the preconditioner is
shown in Table \ref{tab:A2}; in the cut case (iii) the preconditioner is more
efficient than $\mathcal{B}_{2, \Gamma_h}$. We remark that the error convergence
of $\norm{u-u_H}_1$ is reduced in this case due to the fact that the kink
of the solution cannot be captured within the element.

Finally, we extend $\mathcal{A}_{2, \Gamma_h}$ to the formulation with Lagrange
mutliplier on the intersected elements. To this end, let $\mathcal{S}_H$ denote
elements of $\Omega_H$ intersected by $\Gamma_h$. The space $Q_H$ now consists
of piecewise constants on $\mathcal{S}_H$. Following \cite[\S 4.B]{burman2014projection}
we consider problem: Find $u\in V_H$, $p\in  Q_H$ such that
%
\[
\begin{aligned}
  &\int_{\Omega}\nabla u\cdot\nabla v\,\mathrm{d}x + &{\int_{\Gamma}p v\,\mathrm{d}s} &= \int_{\Omega}f v\,\mathrm{d}x\quad\forall v\in V_H,\\
  &{\int_{\Gamma}q u\,\mathrm{d}s} + &-\sum_{F\in\partial\mathcal{S}}\int_{F}H\jump{p}\jump{q} \mathrm{d}s &= \int_{\Gamma}g q\,\mathrm{d}s\quad\forall q\in Q_H,\\
\end{aligned},
\quad\mbox{or equivalently}\quad
\mathcal{A}_{2, \Omega_H}\begin{pmatrix}u\\p\\\end{pmatrix} = L.
\]
%
Here $\partial\mathcal{S}$ denotes interior and Neumann boundary intersecting
facets of $\mathcal{S}$. We remark that the norms for the inf-sup condition (shown
in \cite{burman2014projection} of $\mathcal{A}_{2, \Omega_H}$) are identical to $\mathcal{A}_{2, \Gamma_g}$.

@@

    preconditioned formulations of \eqref{eq:A2}, $H_0=2^{-4}$ and $h=H/3$.
